---
title: "R Notebook"
output: html_notebook
---

Step 1: Load the packages
```{r}
install.packages('tidyverse')
install.packages('skimr')
library(tidyverse) #wrangle data
library(dplyr) #clean data
library(lubridate)  #wrangle date attributes
library(skimr) #get summary data
library(ggplot2) #visualize data
library(readr)
```
```{r}
# Step 2: Prepare the data and combine them in one data frame.

#Combine the data from Jan 2022 to Dec 2022 into one data frame.

bike_data <- rbind(
  read_csv("202201-divvy-tripdata.csv"),
  read_csv("202202-divvy-tripdata.csv"),
  read_csv("202203-divvy-tripdata.csv"),
  read_csv("202204-divvy-tripdata.csv"),
  read_csv("202205-divvy-tripdata.csv"),
  read_csv("202206-divvy-tripdata.csv"))
```
Step 3: Examine the data frame

```{r}
head(bike_data)
dim(bike_data)
colnames(bike_data)
summary(bike_data)
```

4: Check for NA values

```{r}
sum(is.na(bike_data))
```

5: Drop columns (start_lat, start_lng, end_lat, end_lng)

```{r}
bike_data <- bike_data %>% select(-c(start_lat, start_lng, end_lat, end_lng))
```

6: Check data

```{r}
colnames(bike_data)
```

6a: Add column “ride_length"

```{r}
bike_data <- bike_data %>% mutate(ride_length = ended_at - started_at)
```

6b: Add column "Week_day"
```{r}
bike_data <- bike_data %>% mutate(day_of_week = weekdays(as.Date(bike_data$started_at)))

colnames(bike_data)
```
7: Convert ride_length from from seconds into minutes
```{r}
bike_data$ride_length <- as.numeric(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length/60)
head(bike_data)
```
8: Check number of row
```{r}
nrow(bike_data)
```
9: delete data with negative ride length
```{r}
bike_data <- bike_data[bike_data$ride_length>0,]
```
# Check colum "member_casual"

```{r}
table(bike_data$member_casual)
```

# 
```{r}
bike_data <-  bike_data%>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

# Check to make sure the proper number of observations were reassigned
table(bike_data$member_casual)


```

```{r}
# Add columns that list the date, month, day, and year of each ride
# This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level
bike_data$date <- as.Date(bike_data$started_at) #The default format is yyyy-mm-dd
bike_data$month <- format(as.Date(bike_data$date), "%m")
bike_data$day <- format(as.Date(bike_data$date), "%d")
bike_data$year <- format(as.Date(bike_data$date), "%Y")
bike_data$day_of_week <- format(as.Date(bike_data$date), "%A")

```

```{r}
# Add a "ride_length" calculation to all_trips (in seconds)

bike_data$ride_length <- difftime(bike_data$ended_at,bike_data$started_at)

```

```{r}
# Inspect the structure of the columns
str(bike_data)

```
```{r}
# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(bike_data$ride_length)
bike_data$ride_length <- as.numeric(as.character(bike_data$ride_length))
is.numeric(bike_data$ride_length)

```

```{r}
#Step 7: Removing the bad data and do analysis on the ride length.

#check for data with negative ride length
bike_data <- bike_data[bike_data$ride_length>0,]

#check for data with ride length  more than 1 day (86400 seconds or 1440 mins). There is 3484 rides that the rider use the bike for more than 1 day
sum(bike_data$ride_length > 1440)

#check for extreme outliner: 3356649 seconds or 38.85 days for max ride and 1 second for the least ride length
max(bike_data$ride_length)
min(bike_data$ride_length)

#check for mean and median
mean(bike_data$ride_length)
median(bike_data$ride_length)

#Optional: using summary to check min, max, median and mean
summary(bike_data$ride_length)

```

10: CONDUCT DESCRIPTIVE ANALYSIS

```{r}
# Descriptive analysis on ride_length (all figures in seconds)
#straight average (total ride length / rides)
mean(bike_data$ride_length) 
#midpoint number in the ascending array of ride lengths
median(bike_data$ride_length) 
 #longest ride
max(bike_data$ride_length)
#shortest ride
min(bike_data$ride_length) 
#summary
summary(bike_data$ride_length)


```
```{r}
# Compare members and casual users
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = mean)
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = median)
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = max)
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = min)

```

```{r}
# See the average ride time by each day for members vs casual users

aggregate(bike_data$ride_length ~ bike_data$member_casual + bike_data$day_of_week, FUN = mean)

```

```{r}
# Notice that the days of the week are out of order. Let's fix that.
bike_data$day_of_week <- ordered(bike_data$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(bike_data$ride_length ~ bike_data$member_casual + bike_data$day_of_week, FUN = mean)

```

```{r}
# analyze ridership data by type and weekday
#creates weekday field using wday()
bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  
  #groups by usertype and weekday
  summarise(number_of_rides = n()							
  #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		
  # calculates the average duration
  arrange(member_casual, weekday)
  # sorts
```

```{r}
# Let's visualize the number of rides by rider type AND average duration
bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

#Save as csv for further analysis and visualization in Tableau

```{r}
#the data frame

write_csv(bike_data, "bikedata.csv")

#total and average weekly rides by rider type

summary_ride_weekly <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  

write_csv(summary_ride_weekly, "summary_ride_weekly.csv")

#total and average weekly rides by rider type

summary_ride_weekly_type <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  
  
write_csv(summary_ride_weekly_type, "summary_ride_weekly_type.csv")

#total and avg monthly rides by rider type

summary_month <- bike_data %>% 
  mutate(month = month(started_at, label = TRUE)) %>%  
  group_by(month,member_casual) %>%  
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%    
  arrange(month, member_casual)
write_csv(summary_month, "summary_ride_monthly.csv")

#most popular stations

popular_stations <- bike_data %>%
  mutate(station = start_station_name) %>%
  drop_na(start_station_name) %>%
  group_by(start_station_name, member_casual) %>%
  summarise(number_of_rides=n())

write_csv(popular_stations, "popular_stations.csv")

#total membership types and rideable types

total_riders <- data.frame(table(bike_data$member_casual))
total_types <- data.frame(table(bike_data$rideable_type))

write_csv(total_riders, "total_riders.csv")
write_csv(total_types, "total_types.csv")
```
```{r}
# Create number of rides by rider type

number_ride_by_rider <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

ggsave("number_ride_by_rider.png", plot = number_ride_by_rider, width = 6, height = 4, units = "in")

```

```{r}
# Create Average Duration

average_duration <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

ggsave("average_duration.png", plot = average_duration, width = 6, height = 4, units = "in")


```

